CODECHECK: An open-science initiative to facilitate sharing of computer programs and results presented in scientific publications.

### 2020-10-01 codecheck-paper.tex
move overview figure up and add dimensions figure
add text for variations/dimensions

CODECHECK is best demonstrated by way of our example workflow, and later
we expand on the underlying principles. The workflow involves three
groups of people: (1) the AUTHOR of a paper providing the code to be checked, (2)
the PUBLISHER of a journal intereseted in publishing the paper by the
AUTHOR and (3) the CODECHECKER who checks that the AUTHOR's code works.
The workflow that we have refined is documented in Figure 2.

@@

\begin{figure}
% original figure source: https://docs.google.com/drawings/d/1tGOZFNZle-oE1Ynw_tLZFmNcv3wgxD0HPs5bXzDOrng/edit
% emojis: https://github.com/googlefonts/noto-emoji and https://github.com/twitter/twemoji
% emoji codes:
% book stack: 1f4da
% package: 1f4e6
% file drawer: 1f5c4
% disk: 1f4be
% document: 1f4c4
% phd: 1f393
% detective: u1f575
% laptop: 1f4bb
% folder: 1f4c1 1f4c2
% recept: 1f9fe
  \centering
      \includegraphics{figs/codecheck_overview.pdf}
  \caption{The CODECHECK example process implementation. Codechecker act as a detectives:
  The investigate and record, but do not fix issues.}
  \label{fig:worfklow}
\end{figure}

@@

DANIEL: COULD YOU DO THIS SECTION ON DIMENSIONS?
This example workflow leaves sufficient room for different
implementations of this process. There are several aspects that can be
considered:

\subsubsection*{When to do a codecheck}\label{when-to-do-a-codecheck}

preprint (no publisher available), pre-condition for peer review, during
peer review, after acceptance. Each of these have their own pros/cons.

@@

in house staff, pool of ECRs, volunteeres, for inhouse staff, credit not
an issue.

@@

single blind, double blind, private/public communication between author
and codechecker, collaborative. (This might need finessing with
principle 2 below).

>>>>

CODECHECK is best demonstrated by way of our example workflow, and later
we expand on the underlying principles. The workflow involves three
groups of people: (1) the AUTHOR of a paper providing the code to be checked, (2)
the PUBLISHER of a journal intereseted in publishing the paper by the
AUTHOR and (3) the CODECHECKER who checks that the AUTHOR's code works.
The workflow that we have refined is documented in Figure~\ref{fig:workflow}.

% TODO: how is a CODECHECK different from a "specialist reviewer" ? Is it because it is not special, but regular?

\begin{figure}
% original figure source: https://docs.google.com/drawings/d/1tGOZFNZle-oE1Ynw_tLZFmNcv3wgxD0HPs5bXzDOrng/edit
% emojis: https://github.com/googlefonts/noto-emoji and https://github.com/twitter/twemoji
% emoji codes:
% book stack: 1f4da
% package: 1f4e6
% file drawer: 1f5c4
% disk: 1f4be
% document: 1f4c4
% phd: 1f393
% detective: u1f575
% laptop: 1f4bb
% folder: 1f4c1 1f4c2
% recept: 1f9fe
  \centering
      \includegraphics[width=\textwidth]{figs/codecheck_overview.pdf}
  \caption{The CODECHECK example process implementation. Codechecker act as a detectives:
  The investigate and record, but do not fix issues.}
  \label{fig:worfklow}
\end{figure}

@@

@@

\subsubsection*{Dimensions of CODECHECK processes}\label{dimensions-of-processes}

The example workflow leaves sufficient room for different
implementations of this process. To elaborate on these options, we consider
the following dimensions in a space of possible CODECHECKs, as shown in 
Figure~\ref{fig:dimensions}, to represent relevant aspects for the involved
stakeholders. These aspects touch on timing, responsibilities, and 
transparency and will be discussed in the following sections.
Each of these have their own pros and cons and can for the most part be
freely combined.

\begin{figure}
% original figure source: https://docs.google.com/drawings/d/1q2EdW3Ad-IpcD-CeoR1eIFMJW2fF6OVmDhCtCUmLO1k/edit
  \centering
      \includegraphics[width=0.8\textwidth]{figs/codecheck_dimensions.pdf}
  \caption{The dimensions of implementing a CODECHECK process.}
  \label{fig:dimensions}
\end{figure}

\subsubsection*{When to do a codecheck and with what power}
\label{when-to-do-a-codecheck}

The time for a CODECHECK greatly depends on the intentions and the ascribed
importance. The earlier a CODECHECK happens in the process, the more the
results can play a role in the most important decision in scholarly 
communication: is a paper published, sent back for revisions, or rejected.
The later in a review process the check happens, the less challenging is it
to allow a bidirectional communication between author and codechecker, e.g.,
because the author might already be notified of the acceptance and may more
freely share materials online.
Even earlier checks, i.e., a CODECHECK of a preprint, may even help to 
improve the workflow itself even before a publisher is involved. The 
possibility for codechecking papers could be part of a preprint server's
policy, or simply be initiated by interested readers and community members.

A pre-review CODECHECK allows editors to use it as a filter to only send 
a submission out to peer review if it passes the check. A less strict
pre-review check would simply be included in the submission package provided
to the reviewers, so they can include the CODECHECK's results in their 
evaluation of the work.
A CODECHECK may also be conducted in parallel to the scientific review. This
puts less burden on the turnaround time for the check yet only makes the 
outcomes available for the final consideration of the handling editor.
The check could also be tasked on demand, e.g., only if one of the reviewers
suggests/requests one because previous screenings might not have caught the 
fact that a computational workflow is part of the submission. However, 
soliciting such a "specialist review" is much more undesirable than having
a regular CODECHECK, without any special treatment for some submissions.
A CODECHECK could then be awared equal weight as the other reviews.
A post acceptance CODECHECK puts the least requirements on the author and may
simply lead to an award of excellence on top of the acceptance of the
submission. This is the least strict solution while still properly 
acknowleding the check results, because the CODECHECK is to be completed 
before the publication of the final paper.
The GIScience group of checks (see \nameref{register}) falls into this
category: the AGILE conference highlights articles where the reproducibility
was successfully reviewed with a badge on the volume and article landing
pages.

Finally, a CODECHECK may also be conducted after publication, though it
would required to update the article later with a note to reference the check,
otherwise readers would presumably not be aware of it. While this has the
least impact on the current publishing practices, it also has the smallest
effect and does not acknowledge the importance of reproducible workflows for
ensuring good scientific practice.

@@

The options for the person who can conduct a CODECHECK require a not quite
simple matching of a combination of skills and availability. Ideally, the
codechecker has a matching code \emph{and} domain expertise, for example to
check a workflow based on Python code analysing MRT brain imagery. However,
a well-documented workflow should be executable by any person with a basic
understanding of computers. Naturally, the more prerequisite knowledge the
codechecker has, the less time is spend to understand the goals and 
the mechanics of an analysis. From our experiences, the priority should be
given to matching technical expertise first, as the a lack of knowledge in
setting  up a computing environment with a particular language or tool is 
much more discouraging and the assessment of the outcome, e.g., comparing
created figures with the original, is largely possible without 
understanding.
The time and expertise alloted to the check
are the main drivers for the depth of checking, though in general, we 
expect a CODECHECK not to evaluate performance, e.g., does a racing car
have the best apex speed, but to assess operation free of failure, i.e., 
is it a car with 4 wheels and at least one door, and does it move forwards
when accelerating. This depth can be seen independent from the power of
the check (see above), as long as it is communicated clearly to all
stakeholders, most importantly authors and readers.

The concrete people to provide these expertises can range from researchers,
possibly drawn from the regular pool of volunteer reviewers or from a 
special group, to editorial staff of a publisher. If the regular reviewers
are relied upon, the editor's task to find suitable ones becomes more
tedious; extra information such as software familiarity would have to be
modelled in review platforms.
We see a great opportunity to especially involve early career 
researchers as codecheckers. They arguably have a high interest in learning
about the latest tools and technologies as they are building up their own
expertise and specialisation. They are also introduced to peer review and 
may transition into the role of scientific reviewer over time.
Not to solve the general problems of peer review, which is largely an
unsupervised process, i.e., rarely are reviewers taught how to do it, but
a junior codechecker may be supported by a senior codechecker, too.
We see a high potential in setting up the new processes for CODECHECKING with
a clear committment on opennes and transparency, independent of the
respective current peer review process (see \nameref{who-knows-who}).

Publisher staff represents the most controlled yet also not quickly scalable
option. Hiring the required technical and or domain expertise puts a financial
burden on the publisher, of course, but such a committment also shows the 
value given to publishing reproducibility and would positively impact the 
submission experience, because of more controllability, and external 
perception of the journal or conference. A staff codechecker reduces the 
challenges when it comes to open communication and awarding credit, as
in-house personnel may be more readily be given access to non-anonymised
submission material and receives a proper salary instead of public credit
as part of the scientific community.

In contrast, for researchers it can be very important to be publicly credited
with their activity as a reviewer (see \nameref{who-knows-who}). A regular
review may be listed in public databases (e.g., ORCID
\footnote{\href{https://support.orcid.org/hc/en-us/articles/360006971333-Peer-Review}{https://support.orcid.org/hc/en-us/articles/360006971333-Peer-Review}} or commercial offerings such as Publons
\footnote{\href{https://publons.com/}{https://publons.com/}} or 
ReviewerCredits\footnote{\href{https://www.reviewercredits.com/}{https://www.reviewercredits.com/}}).
A codechecker should be listed just the same. The number of scientific reviews
if of course a very coarse indicator for the community contribution by an 
individual researcher, so it's unclear if a regular reviewer, who in addition
conducts a CODECHECK to the scientific review, should be credited twice.

The codecheckers community\footnote{\href{https://github.com/codecheckers/codecheckers/}{https://github.com/codecheckers/codecheckers/}}
currently has over 20 individuals who signed up as volunteer codecheckers
following talks given on the project in the last 12 months.
That is even though no checks are actively elicited.
We see a high potential in an open shared list of potential codecheckers 
compared to a private, in-house group, as it may better match expertise
and share the workload.
% TODO tell more about the motivations and commitments that we know about?

@@

The topic of blinding is broadly discussed, not the least in the push towards
open peer review as part of the Open Science movement [REF]. Without taking
a strong stand on this discussion, the general motivation for more 
transparency and reproducibility does indeed favour a more open review 
process. However, anonymity can also serve to protect individuals 
\ref{tennant_limitations_2020}, e.g.,  juniour scientists, whereas it might 
be acceptable to require revealing of the names of tenured researchers.
The possible negative effects of a non-anonymous review are greatly reduced
if a CODECHECK is not relevant for the decision to accept or reject, but that
is of course not desirable. Instead, we argue that a CODECHECK is technical
process that should in any case be fixable and not a question of opinion or
faulty approach. It is possible for a nonsense workflow to receive a 
CODECHECK certificate.
Furthermore, the level of documentation that is needed for third parties
to reproduce a workflow is extremely hard to get right. Too often, this 
uncertainty leads to researchers giving up and not documenting at all.

The technical nature of the check and the challenge of good enough 
documentation is why we see great benefits in a effective bidirectional
means of communication between author and codechecker. Instead of trying to
fix problems or guess the next step, the codechecker can ask the author to 
rework the docs  or the code.
Instead of lenghtily describing a small change, the codechecker may simply
provide a fix within the code.
Instead of hiding useful and  instructive files created during the codecheck
(e.g., a machine-readable computing environment specification), the author 
and readers can profit from the software-related expertise of the
codechecker.
While such a communication may be facilitated in an anonymous way by the 
publisher, it most likely only helps to protect the identity of the 
codechecker, because, especially when of a certain size, not trivial, or 
if very specialised, it is close to impossible to fully anonymise a 
software analysis stack.
Therefore, the most effective means with the most desirable result for
the stakeholders is a non-anonymous and collaborative CODECHECK.
The possible contributions by the codechecker may even be integrated into
the code of the workflow and be acknowledged as code commits. This way, 
proper credit is given within the research software development community.
Nevertheless, depending on other needs represented by other dimensions,
the aspect of `Who knows who?' may be the most flexible one.

\subsubsection*{How is the CODECHECK integrated in other processes?}\label{turnaround-time}

As a prerequisite, we see a CODECHECK as part of a typical scholarly
communication process. While more extreme ideas to change the paradigms
how researchers share their work exist [REF Octopus], there is a 
value in an evolutionary approach. Thereby, existing communities can 
transition towards more open practices at their own pace and ensure
no one is left behind unfairly.
When integrating a CODECHECK in existing review and publication processes,
the \emph{turnaround time} is crucial. Depending on when and who conducts
the check, it may be done faster, or add considerably to the overall
time from submission to publication.
% TODO write about our experiences how long a CODECHECK takes?
